Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/elementary.tex
fixed captitalization in captions

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Cycles per key - this was measured on a Pentium 4 using hardware performance counters.]
{\label{elementary cycles}\includegraphics[scale=\myscale]{plots/on2_-_cycles.eps}}
\subfigure[Instructions per key - this was simulated using SimpleScalar \cc{sim-cache}.]	
{\label{elementary instructions}\includegraphics[scale=\myscale]{plots/on2_-_Instructions.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulated Instruction count and empiric cycle count for elementary sorts}
\label{Simulated Instruction count and empiric cycle count for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
cache with a 32-byte cache line and separate instruction cache.]
{\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}.]
{\label{elementary branches}\includegraphics[scale=\myscale]{plots/on2_-_Branches.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Cache and Branch prediction simulation results for elementary sorts}
\label{Cache and Branch prediction simulation results for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\centering
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses1.eps}}
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses2.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Branch Simulation results for elementary sorts}
\label{Branch Simulation results for elementary sorts}
\end{figure}
}

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key. Due to huge variations in predictability, we split
the numbers of branch mispredictions into two charts. Selection sort causes
around $H_N$ mispredictions per key, which is remarkably small. The comparison
branch in the inner loop is highly predictable on random data, because a good
candidate minimum is usually found quickly.

>>>>

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Cycles per key - this was measured on a Pentium 4 using hardware performance counters.]
{\label{elementary cycles}\includegraphics[scale=\myscale]{plots/on2_-_cycles.eps}}
\subfigure[Instructions per key - this was simulated using SimpleScalar \cc{sim-cache}.]	
{\label{elementary instructions}\includegraphics[scale=\myscale]{plots/on2_-_Instructions.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Simulated instruction count and empiric cycle count for elementary sorts}
\label{Simulated instruction count and empiric cycle count for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\subfigure[Level 1 cache misses per key - this was simulated using SimpleScalar \cc{sim-cache}, simulating an 8KB data
cache with a 32-byte cache line and separate instruction cache.]
{\label{elementary level 1 misses}\includegraphics[scale=\myscale]{plots/on2_-_Level_1_misses.eps}}
\subfigure[Branches per key - this was simulated using \cc{sim-bpred}.]
{\label{elementary branches}\includegraphics[scale=\myscale]{plots/on2_-_Branches.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Cache and branch prediction simulation results for elementary sorts}
\label{Cache and branch prediction simulation results for elementary sorts}
\end{figure}
\newpage

\thispagestyle{empty}
\clearpage
\enlargethispage{14em}
\vspace*{-8em}
\begin{figure}[H]
\begin{changemargin}
\centering
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses1.eps}}
\subfigure[Branches misses per key - this was simulated using \cc{sim-bpred}, with bimodal and two-level adaptive
predictors. The simulated two-level adaptive predictor used a 10 bit branch history register which was XOR-ed with the
program counter.]
{\includegraphics[scale=\myscale]{plots/on2_-_Branch_misses2.eps}}
\end{changemargin}
\vspace*{6em}
\caption{Branch simulation results for elementary sorts}
\label{Branch simulation results for elementary sorts}
\end{figure}
}

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch simulation results for elementary sorts} shows
the branch misses per key. Due to huge variations in predictability, we split
the numbers of branch mispredictions into two charts. Selection sort causes
around $H_N$ mispredictions per key, which is remarkably small. The comparison
branch in the inner loop is highly predictable on random data, because a good
candidate minimum is usually found quickly.

### 2005-08-14 docs/tech_report/heap.tex
fixed usage of '32 byte' without a hyphen
fixed captitalization in captions

\section{Base Heapsort}
LaMarca recommends to use of the Repeated-Adds algorithm for reducing
cache misses. As such, his base heapsort uses Floyd's method, ostensibly at
the behest of major textbooks, each of whom claim Floyd's method gives better
results. This is certainly backed up by Sedgewick, who recommends this method,
but in initial tests, the out-of-place Repeated-Adds technique had a lower
instruction count. Regardless, the base algorithm used here is Floyd's in-place
sort. Code for a base heapsort is in Figure \ref{Simple Base Heapsort}.

\code{Simple Base Heapsort}{base_heapsort.c}

@@

\section{Results}
All simulations presented here were performed on 32-byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \cc{int}s fit the cache
line.  LaMarca used 64 bit ints, and while this doesn't affect results, it means
that fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 ints, we stick with 32 byte cache lines. \cc{UINT\_MAX} was used
as the sentinel.

>>>>

\section{Base Heapsort}
LaMarca recommends to use of the Repeated-Adds algorithm for reducing
cache misses. As such, his base heapsort uses Floyd's method, ostensibly at
the behest of major textbooks, each of whom claim Floyd's method gives better
results. This is certainly backed up by Sedgewick, who recommends this method,
but in initial tests, the out-of-place Repeated-Adds technique had a lower
instruction count. Regardless, the base algorithm used here is Floyd's in-place
sort. Code for a base heapsort is in Figure \ref{Simple base heapsort}.

\code{Simple base heapsort}{base_heapsort.c}

@@

\section{Results}
All simulations presented here were performed on 32-byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \cc{int}s fit the cache
line.  LaMarca used 64 bit ints, and while this doesn't affect results, it means
that fewer keys fit into his cache lines. Since modern cache lines are typically
longer than 4 ints, we stick with 32-byte cache lines. \cc{UINT\_MAX} was used
as the sentinel.

### 2005-08-14 docs/tech_report/merge.tex
fixed all references
fixed use of =comparison-based (no hyphen) and fixed up nearly all \n vs \cc usage
fixed captitalization in captions

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
This shows there must be regular access patterns in the k-way merge at the end
of multi-mergesort, which are predictable using a branch history.

@@

\section{Predictor Results}

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
heapsort curve}). Note that \n{comparison 0} occurs every time, and would be
optimized out by the compiler.

>>>>

The prediction results in Figure \ref{Branch simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
This shows there must be regular access patterns in the k-way merge at the end
of multi-mergesort, which are predictable using a branch history.

@@

\section{A More Detailed Look at Branch Prediction}

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned
heapsort curve}). Note that `\n{comparison 0}' occurs every time, and would be
optimized out by the compiler.
